---
title: "Bike Data"
author: "MaxClef"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### How do annual members and casual riders use Cyclistic bikes differently? 


Importing libraries
```{r, echo=FALSE,message=FALSE}
library(RPostgres)
library(DBI)
library(getPass)
library(tidyverse)
library(ggplot2)
library(dplyr)
```

**Based on all the information found while working with the data in SQL, I decided to import:**

* data that includes only `classic_bike` and `electric_bike`
 

Connecting to the database/getting the data
```{r}
con <- dbConnect(RPostgres::Postgres(),
                 dbname = 'divvy-tripdata', 
                 host = "localhost", 
                 port = 5432, 
                 user = "postgres",
                 password = getPass())

query = "
  SELECT 
    m.member_type,
    b.bike_type,
    s1.station_name AS start_station_name,
    s2.station_name AS end_station_name,
    t.duration_minutes,
    t.started_at AS started_time,
    t.ended_at AS ended_time,
    t.distance_meters,
    t.start_lat,
    t.start_lng,
    t.end_lat,
    t.end_lng
  FROM
    trip_data_past_year t
      JOIN member_types m
          ON t.member_type_id = m.id
      JOIN bikes b
          ON t.bike_type_id = b.id
      JOIN station_names s1
          ON t.start_station_name_id = s1.id
      JOIN station_names s2
          ON t.end_station_name_id = s2.id
  WHERE b.bike_type IN ('classic_bike', 'electric_bike');
"

data <- dbGetQuery(con, query)

dbDisconnect(con)
```

Checking the retrieved data
```{r}
str(data)
```

Splitting columns `started_time` and `ended_time` into:

* year - for year (e.g., 2022)

* month - for month (e.g., August)

* day - for day of the week (e.g., Monday)


```{r}
data$year <- format(data$started_time, "%Y")

data$month <- format(data$started_time, "%B")

data$day <- format(data$started_time, "%A")
```


Converting `month` and `day` into factors (when plotting they will be orgered correctlly)  
```{r}
data$month <- factor(data$month,
                     levels = c("October", "November", "December", "January", 
                                "February", "March", "April", "May", "June",
                                "July", "August", "September"))
data$day <- factor(data$day, 
                   levels = c("Monday", "Tuesday", "Wednesday", "Thursday",
                              "Friday", "Saturday", "Sunday"))
```

Checking mutated datarfame
```{r}
str(data)
```


Checking the number of missing values in each column

```{r}
colSums(is.na(data))
```

No missing values are present


### Checking for Inconsitencies 


Visualizing `duration_minuts`
```{r}
ggplot(data) +
  geom_boxplot(aes(x = duration_minutes, fill = "#0c4c8a")) +
  labs( title = "Bike Usage Time", x = "Minutes") +
  theme( legend.position = "none")
```

 Percentiles 
 
```{r}
quantile(data$duration_minutes)
```

Descriptive statistics for `duration_minuts`

```{r}
psych::describe(data$duration_minutes)
```

Looking at the maximum duration for 95% of the time

```{r}
upper_bound <- quantile(data$duration_minutes, 0.95)
upper_bound
```

It's clear that variable `duration_minutes` is right skewed based on the mean (16.26) and
trimmed mean (12.49)

I doubt there are people that use the bikes for up to 1000+ minutes (16+ hours) 

Even if they do these are clear outliers, I mean the pricing is steep:

* members :

  * classic bikes ($119 per year):
  
    - unlocking the bike - free
    - usage - free 45 minutes at a time
    - over 45 minutes is $0.16 per minute
    
  * electric bikes ($119 per year)
  
    -unlocking the bike - free
    - usage - $0.16 per minute
    - parking has extra charges based on location
    
* casual:

  * classic bikes:

    - unlocking the bike - $1
    - usage - $0.16 per minute

  * electric bikes 
  
    -unlocking the bike - $1
    - usage - $0.39 per minute
    - parking has extra charges based on location
    
There is also a third category of users which are casual members
that get a `daily pass`. They have 3 hours of free rides for 24 hours for 
classic bikes.

Sources [here](https://divvybikes.com/pricing) and [here](https://divvybikes.com/pricing/ebikes)

The battery of an electric bike lasts maximum for 60 miles.
If we assume that the person rides one non-stop at a maximum speed 
of 3 hours, thay have enough battery only for 3 hours.

Source [here](https://ride.divvybikes.com/blog/new-divvy-ebike#:~:text=Longer%20rides%3A%20We've%20increased,operations%20trips%20to%20replace%20batteries.)

Based on all of the above I decided to remove all that have a duration
above 43.72 minutes. Meaning, removing 5% (208,951 values) from the data.

```{r}
clean_data <- data %>% filter(duration_minutes <= 43.72 )

```


Visualizing cleaned data

```{r}
ggplot(clean_data) +
  geom_boxplot(aes(x = duration_minutes, fill = "#0c4c8a")) +
  labs( title = "Bike Usage Time", x = "Minutes",
        subtitle = "Cleaned Data") +
  theme( legend.position = "none")
```


### 1. How do annual members and casual riders use Cyclistic bikes differently?


<span style="color:blue">**Difference in time usage by group**</span>

Visualizing duration by group and bike type

```{r}
ggplot(clean_data, aes(x = member_type, y = duration_minutes, fill = member_type)) +
  stat_boxplot(geom = "errorbar", width = 0.25) +
  geom_boxplot() + 
  labs(title = "Bike Usage", subtitle = "Bike Type and User Type", caption = "Past 12 mth",
      x = "Type of User", y =  "Duration") + 
  theme(legend.position = "none") +
  facet_wrap(~bike_type)
```



Mean for each group by usage of bike type

```{r}
clean_data %>% group_by(member_type, bike_type) %>% summarise_at(vars(duration_minutes), 
                                                           list(mean = "mean"))
```

Looks like on average casual users tend to use the bikes for a longer period of time. 


<span style="color:blue">**Bike usage by month for each group**</span>


Plot the difference in usage of bikes by month

```{r}
ggplot(clean_data) +
  geom_bar(aes(x = month, fill = member_type, color = member_type), position = position_dodge(width=0.5)) +
  coord_flip() +
  labs(title = "Bikes Usage", subtitle = "Bike Type and User Type", caption = "Past 12 mth") +
  theme(axis.text.x = element_text(angle=25)) +
  scale_y_continuous(labels = function(x) format(x, scientific = FALSE)) +
  facet_wrap(~bike_type)
```

Based on the plot above we see that the only time when the casual users
use the services more than the members is in the month of July
for electric bikes.Potentially because of summer tourists. 

Another factor for bike usages could be the weather. 

Casual users contingency table by month and type of bike

```{r}
casual_users <- data %>% filter(member_type == "casual")
c_table <- prop.table(table(casual_users$bike_type, casual_users$month)) * 100
addmargins(round(c_table, 2))
```

casual users use classic bikes through out the year more often than electric bikes:

* 58.18% using classical bikes usage
* 41.81% using electric bikes usage

Members contingency table by month and type of bike

```{r}
casual_users <- data %>% filter(member_type == "member")
c_table <- prop.table(table(casual_users$bike_type, casual_users$month)) * 100
addmargins(round(c_table, 2))
```
Members are using classic bike though out the year more often than electric once:

* 67.65% classic bikes usage
* 32.33% electric bikes usage


<span style="color:blue"> **Busiest days of each month**</span>

**Casual**

```{r}

casual <- filter(data, member_type == "casual")
ct_casual <- as.data.frame(table(casual$day, casual$month))

ggplot(ct_casual, aes(x = Var1, y = Var2)) +
  geom_tile(aes(fill = Freq)) + 
  geom_text(aes(label = round(Freq, 4)), size = 2) +
  scale_fill_gradient(low = "white", high = "orange") +
  labs(title = "Usage by Month and Day for Casual Members", x = "Day", y =  "Month") +
  theme(text = element_text(size = 10))
```

Casual members have a higher frequency of bike riding on the weekends




**Mabers**

```{r}

member <- filter(data, member_type == "member")
ct_members <- as.data.frame(table(member$day, member$month))

ggplot(ct_members, aes(x = Var1, y = Var2)) +
  geom_tile(aes(fill = Freq)) + 
  geom_text(aes(label = round(Freq, 4)), size = 2) +
  scale_fill_gradient(low = "white", high = "orange") +
  labs(title = "Usage by Month and Day for Monthly Members", x = "Day", y =  "Month") +
  theme(text = element_text(size = 10))
```

Members seem to be biking more often on the week days.


<span style="color:blue"> **Top 10 popular starting points by group**</span>

**Casual**

```{r}
count(casual, start_station_name, sort = T)[1:10, ]
```

**Members**

```{r}
count(member, start_station_name, sort = T)[1:10, ]
```